A Study of Institutional Spending on Open Access Publication Fees in Germany

### 2016-06-10 article.Rmd
proofreading introduction finished

---
title: 'A Study of Institutional Spending on Open Access Publication Fees in Germany'
author:
- Najko Jahn, Bielefeld University Library, Bielefeld University (TIB) - German National
  Library of Science and Technology, Hannover, Germany
- <a href="http://orcid.org/0000-0002-5111-2788">Marco Tullney</a>, Technische Informationsbibliothek
date: "7 April 2016"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: peerj.sty
  html_document:
    fig_caption: yes
    force_captions: yes
    number_sections: no
    theme: readable
bibliography: literatur.bib
csl: frontiers.csl

@@

The rise of open access journals matches the increasing relevance of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. To cover these fees, also referred to as article-processing charges (APC), authors tend to make use of funding that grant agencies or academic institutions provide [@Suber_2012]. Yet, how and to what extent these research support activities are effective in terms of the number of supported articles and associated costs remains under debate.

The study of institutional spending on open access journal articles has been limited for several reasons. One is that payment on these charges is fragmented across the budgets of grant agencies, research institutions, and libraries, or is covered by personal budgets. Asking 9,645 authors from various disciplines how they financed publication fees, a comprehensive survey in 2010 revealed that the majority of the respondents had access to research funding or institutional support to cover these charges. By contrast, 12 % paid publication fees individually [@soap]. These findings are consistent with that of other studies, adding that sources of funding mostly exists in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets, on the other hand, are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015].

Another key problem in this regard is that funding for open access journals using publication fees lacks transparency because the parties involved - authors, universities, funders, publishers - neither release information on who pays for what nor the costs of publishing [@Bj_rk_learned_2014], a situation similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015]. Empirical studies examining publication fees so far obtained price estimates by surveying authors [@soap], or from journal websites. Using the latter method, two studies investigating journals across a broad range of disciplines calculated similar price averages that ranged between 904 $ [@Solomon_2012] and 923 $ [@Walters_2011], as well as considerable price variances across journals and publishers. Accordingly, @Solomon_2012 suggested to cluster fully open access journal using publication fees into several groups. In descending order, these are high-impact journals, followed by biomedicine journals from commercial publishers, large multi-disciplinary journals, and mid-price journals from commercial publishers covering a large spectrum of disciplines. Lower priced journals were published by academic societies and by publishers from low-income countries.

@@

Hybrid journals substantially add to this complexity of open access funding [@Bj_rk_learned_2014; @Kingsley_2014; @Pinfield_2015]. These journals, allowing articles to be published immediately as open access after a charge was paid, rely both on subscriptions and publication fees as revenue sources. Although the uptake of open access through hybrid journals was described as lower and more expensive compared to that of fully open access journals [@Solomon_2012; @Bj_rk_learned_2014], this model has gained increasing attention because of recent open access science policies, notably from the UK [@Pinfield_Review_2015].

To address these problems of fragmented spending on publication fees and of in-transparency about what was being paid, some European research funders and research performing institutions have recently begun to disclose their payment for publication as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc_12] and the Austrian Science Fund FWF [@fwf_apc_13]. The not-for-profit company Jisc followed by collecting data from UK universities [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report expenditure along with bibliographic information, including title, journal and publisher, and a persistent identifier to the publisher's version. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates and persistent identifiers to the full text including the Digital Object Identifier (DOI) [@Neylon_2014; @woodward_2014]. Parts of Jisc's cost data was examined by @Pinfield_2015. Although the average spending on publication fees remained stable across the universities, they found large price variances, as well as a varying number of articles UK universities supported between 2007 - 2014, being consistent with earlier studies that collected price information from journal websites [@Solomon_2012].

@@

This paper focuses on how much German universities and research organisations spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced how universities manage institutional support for these charges. Before the DFG started to pay for centrally funded publication fees on a pro rata basis through its "Open-Access Publishing" program in 2011, only a few central funds existed [@eppelin_2012]. This is similar to the situation described in Canada [@Hampson_2014] or the UK [@PINFIELD_2012]. The DFG has enforced a set of criteria grantees have to comply with  [@Fournier_2013], leading to related implementations for open access supporting across German universities. These criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 €[^3]. Grantees agree not only to pay for APCs, but also to find ways to improve the handling of those financial transactions. This includes central invoicing schemes and memberships, among others, agreed between university libraries and publishers that often lead to a discount on publication fees [@Fournier_2013].

Non-university research organisations, i.e. institutes organised in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft, are not eligible for this DFG funding program. But in response, some have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates their long-lasting open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library (MPDL) [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

The evolving institutional support to cover open access publication fees has led to calls for an unified approach towards open access funding in Germany. The MPDL called for re-allocating subscriptions in favour of open access  journals in 2015 [@Schimmer_Geschuhn_Vogler]. At the same time, the Allianz der Wissenschaftsorganisationen[^10c], a science policy board representing all major research organisations in Germany, marked price transparency as one way to sustain an "adequate open access publication system" [@allianz]. Reflecting Austrian and UK initiatives to share institutional spending on open access publication fees as open data, as well as professional discussions on open access publishing, Bielefeld University Library began to openly share its payment of publication fees in May 2014. After engaging with the working group "Electronic Publishing" of the Deutsche Initiative für Netzwerkinformation (DINI)[^11], other German institutions joined under the umbrella of the Open APC initiative soon after [@openapc].

@@

The aim of the study was to examine how much German universities and research organisations spent on open access publication fees until 2015. Using self-reported cost data from the Open APC initiative, the analysis focused on the amount that was being spent on publication fees, and compared these expenditure with data from related Austrian and UK initiatives, both in terms of size and the proportion of articles being published in fully and hybrid open access journals. We also asked how thoroughly self-reported articles were indexed in Crossref, a DOI minting agency for scholarly literature, and analysed how the institutional spending was distributed over publishers and journal titles.

>>>>

---
title: 'A Study of Institutional Spending on Open Access Publication Fees in Germany'
author:
- Najko Jahn, Bielefeld University Library, Bielefeld University 
- <a href="http://orcid.org/0000-0002-5111-2788">Marco Tullney</a>, TIB - German National Library of Science and Technology, Hannover, Germany
date: "7 April 2016"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: peerj.sty
  html_document:
    fig_caption: yes
    force_captions: yes
    number_sections: no
    theme: readable
bibliography: literatur.bib
csl: frontiers.csl

@@

The rise of open access journals has been matched by the increasing relevance of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. To cover these fees, also referred to as article-processing charges (APCs), authors tend to make use of funding that grant agencies or academic institutions provide [@Suber_2012]. However, the question of how and to what extent these research support activities are effective in terms of the number of supported articles and their associated costs remains under debate.

The study of institutional spending on open access journal articles has been limited for several reasons. The first is that payment of these charges is fragmented across the budgets of grant agencies, research institutions, and libraries, or is covered by personal budgets. A comprehensive 2015 survey asked 9,645 authors from various disciplines how they financed publication fees, and it revealed that the majority of the respondents had access to research funding or institutional support to cover these charges. By contrast, 12 % paid publication fees individually [@soap]. These results are consistent with similar findings from other studies. Previous studies also found that funding sources exist primarily in higher-income countries, mainly to support research articles in the biological and physical sciences [@Solomon_2011]. Personal budgets, however, are likely used to cover lower publication fees [@Solomon_2011; @Bj_rk_2015].

Another key problem in this regard is that funding for open access journals using publication fees lacks transparency because the parties involved -- authors, universities, funders, and publishers -- do not  release information about who pays for what or the costs of publishing [@Bj_rk_learned_2014]; this situation is similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015]. To date, empirical studies examining publication fees have obtained price estimates by surveying authors [@soap] or obtained them from journal websites. Using the latter method, two studies investigating journals across a broad range of disciplines calculated similar price averages that ranged between $ 904 [@Solomon_2012] and $ 923 [@Walters_2011], as well as considerable price variation across journals and publishers. Accordingly, @Solomon_2012 suggested using publication fees to cluster fully open access journal into several groups. In descending order, these are high-impact journals, followed by biomedicine journals from commercial publishers, large multi-disciplinary journals, and mid-price journals from commercial publishers covering a large spectrum of disciplines. Lower-priced journals are those published by academic societies and by publishers from low-income countries.

@@

Hybrid journals substantially add to the complexity of open access funding [@Bj_rk_learned_2014; @Kingsley_2014; @Pinfield_2015]. These journals, which allow articles to be published immediately as open access after a charge is paid, rely on both subscriptions and publication fees as revenue sources. Although the uptake of open access through hybrid journals was described as lower and more expensive than that of fully open access journals [@Solomon_2012; @Bj_rk_learned_2014], this model has gained increasing attention because of recent open access science policies, notably from the UK [@Pinfield_Review_2015].

To address the problems of fragmented spending on publication fees and the lack of transparency about what is being paid, some European research funders and research-performing institutions have recently begun to disclose their expenditures for publication as open data. To the best of our knowledge, the first research funders to provide such data were the Wellcome Trust [@wellcome_apc_12] and the Austrian Science Fund FWF [@fwf_apc_13]. The not-for-profit company Jisc followed this example by collecting data from UK universities [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report expenditures along with bibliographic information, including title, journal and publisher, and a persistent identifier for the publisher's version. Curatorial efforts focussed on the disambiguation of publisher and journal titles as well as on detecting duplicates and persistent identifiers to the full text including the Digital Object Identifier (DOI) [@Neylon_2014; @woodward_2014]. A preliminary version of Jisc's cost data was examined by @Pinfield_2015. Although the average spending on publication fees remained stable across universities, the authors found large price variations, as well as a varying number of articles supported by UK universities between 2007 and 2014, findings which confirm earlier studies that collected price information from journal websites [@Solomon_2012].

@@

This paper focuses on how much German universities and research organisations spend on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced how universities manage institutional support for these charges. Before the DFG started to pay for centrally funded publication fees on a pro rata basis through its "Open-Access Publishing" programme in 2011, only a few central funds existed [@eppelin_2012]. This is similar to the situation described in Canada [@Hampson_2014] and the UK [@PINFIELD_2012]. The DFG has enforced a set of criteria with which grantees have to comply and which has resulted in similar policies regarding support for publication fees across German universities [@Fournier_2013]. These criteria exclude the sponsorship of articles in hybrid journals and the funding of articles for which the publication fee exceeds € 2,000[^3]. Grantees agree not only to pay for APCs, but also to find ways to improve the handling of those financial transactions. These ways include central invoicing schemes and memberships that are agreed upon by university libraries and publishers and that often lead to a discount on publication fees [@Fournier_2013].

Non-university research organisations, i.e., institutes organised in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft, are not eligible for this DFG funding programme. However, in response, some organisations have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates its long-standing open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library (MPDL) [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

The evolving institutional support for covering open access publication fees has led to calls for a unified approach towards open access funding in Germany. The MPDL called for re-allocating subscriptions in favour of open access  journals in 2015 [@Schimmer_Geschuhn_Vogler]. At the same time, the Allianz der Wissenschaftsorganisationen[^10c], a science policy board representing all major research organisations in Germany, marked price transparency as one way to sustain an "adequate open access publication system" [@allianz]. Reflecting Austrian and UK initiatives to share institutional spending on open access publication fees as open data, as well as professional discussions on open access publishing, Bielefeld University Library began to openly share its payment of publication fees in May 2014. After engaging with the working group "Electronic Publishing" of the Deutsche Initiative für Netzwerkinformation (DINI)[^11], other German institutions joined under the umbrella of the Open APC initiative soon after [@openapc].

@@

The aim of this study was to examine how much German universities and research organisations spent on open access publication fees. Using self-reported cost data from the Open APC initiative, the analysis focussed on the amount that was being spent on publication fees, and compared these expenditure with data from related Austrian and UK initiatives, in terms  of both size and the proportion of articles being published in fully and hybrid open access journals. We also investigated how thoroughly self-reported articles were indexed in Crossref, a DOI minting agency for scholarly literature, and analysed how the institutional spending was distributed across publishers and journal titles.
